---
title: "Buğra ATEŞ"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Buğra ATEŞ B1705.090046

```{r}
library(limma)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
 

 
covid_dataset_df <- read.csv("C:/Users/bugra/Desktop/Buğra ATEŞ R Project/covid19-GSE160435")
#reading dataset and save to covid_dataset as 

```


```{r}
# --- 1.apply normalization ---

covid_dataset_class <- covid_dataset_df[,1]
#seperating class column to another variable

covid_no_normalize <- covid_dataset_df[, 2:dim(covid_dataset_df)[2]]
# saving dataset with no "class" column for normalizing

covid_normalized <- normalizeBetweenArrays(covid_no_normalize, method = "quantile")
# normalizing dataset with "quantile" method in "limma" package
```


# --- 2.draw boxplots before and after normalization ---
```{r}
boxplot(covid_no_normalize[,1:30])
# showing unnormalized dataset with boxplot
```

```{r}
boxplot(covid_normalized[,1:30])
# showing normalized dataset with boxplot
```


# --- 3. find statistically differential expression genes using t-test. ---
```{r}
t.test(covid_normalized[,1],covid_normalized[,3],mu=0, alt="two.sided", conf=0.95, var.eq=F, paired=F)
# t-test between virus_N and virus_M
```

```{r}
t.test(covid_normalized[,3],covid_normalized[,5],mu=0, alt="two.sided", conf=0.95, var.eq=F, paired=F)
# t-test between virus_M and virus_S
```

```{r}
t.test(covid_normalized[,7],covid_normalized[,8],mu=0, alt="two.sided", conf=0.95, var.eq=F, paired=F)
# t-test between virus_ORF7a and virus_ORF3a
```

```{r}
t.test(covid_normalized[,8],covid_normalized[,9],mu=0, alt="two.sided", conf=0.95, var.eq=F, paired=F)
# t-test between virus_1_261 and virus_ORF10
```

```{r}
t.test(covid_normalized[,2],covid_normalized[,6],mu=0, alt="two.sided", conf=0.95, var.eq=F, paired=F)
# t-test between virus_ORF1ab and virus_ORF7a
```


# --- 4.show 100 top genes based on adjusted-pvalues ---
```{r}
class_factor <- as.factor(covid_dataset_class)
# encoding vector as a factor
```

```{r}
class_numeric<- as.numeric(class_factor)
# encoding factor as a numeric
```

```{r}
dataset_fit <- lmFit(t(covid_no_normalize), class_numeric)
# fit dataset - t() for replacing places between columns and rows
```

```{r}
bayes_fit <- eBayes(dataset_fit)
# using eBayes method
```

```{r}
adjusted_p_values<- topTable(bayes_fit, adjust.method = 'BH', number=100, sort.by = 'p');
# creating top table based on adjusted p-values
```

```{r}
column_names <-row.names(adjusted_p_values)[1:100]
# taking column names with row.names() function because we replaced the places 
```

```{r}
adjusted_p_values_top_100 <- covid_normalized[, column_names]
# declaring variable with top 100 genes based on adj.p-values and combining with their column names
```

# ---5. draw heatmap and hierarchical clustering for extracted 100 top genes ---
```{r}
adjusted_p_values_top_100_matrix <- t(adjusted_p_values_top_100)
# replacing column and rows again
```

```{r}
colnames(adjusted_p_values_top_100_matrix) <- covid_dataset_class
# replacing V1-V2....V10 columns with "covid_dataset_class" column that we declared in the beginning
```

```{r}
adjusted_p_values_top_100_matrix <- as.matrix(adjusted_p_values_top_100_matrix)
# converting to matrix to use heatmap()
```

#----------------------HEATMAP-------------------------
```{r}
heatmap_color <- colorRampPalette(brewer.pal(10, "Blues"))(25)
# declaring color for heatmap

heatmap(adjusted_p_values_top_100_matrix , col=heatmap_color )
# executing heatmap() function to create heatmap for extracted 100 top genes
```

#---------------HIERARCHICAL CLUSTERING-----------------

```{r}
top_100_clusters <- hclust(dist(adjusted_p_values_top_100_matrix)) 
# doing hierarchical clustering wth hclust() function

plot(top_100_clusters) 
# visualizing cluster dendrogam plot()
```


# --------- 6.Do any other meaningful statistical analysis. ------------

#-----------------GEOM BOXPLOT BEFORE NORMALIZATION-------------------------------
```{r}
ggplot(covid_dataset_df, aes(x=class, y=virus_N, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_dataset_df, aes(x=class, y=virus_ORF1ab, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_dataset_df, aes(x=class, y=virus_M, fill=class)) + geom_boxplot
```

```{r}
ggplot(covid_dataset_df, aes(x=class, y=virus_ORF8, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_dataset_df, aes(x=class, y=virus_S, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_dataset_df, aes(x=class, y=virus_ORF7a, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_dataset_df, aes(x=class, y=virus_ORF3a, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_dataset_df, aes(x=class, y=virus_1_261, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_dataset_df, aes(x=class, y=virus_ORF10, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_dataset_df, aes(x=class, y=MMP9, fill=class)) + geom_boxplot()
```


#-----------------GEOM BOXPLOT AFTER NORMALIZATION-------------------------------
```{r}
class <- covid_dataset_df[,1] 
# classes for ggplot

covid_class_with_normalized <- data.frame(class,covid_normalized)
# combining class and normalized data for ggplot
```

```{r}
ggplot(covid_class_with_normalized, aes(x=class, y=virus_N, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_class_with_normalized, aes(x=class, y=virus_ORF1ab, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_class_with_normalized, aes(x=class, y=virus_M, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_class_with_normalized, aes(x=class, y=virus_ORF8, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_class_with_normalized, aes(x=class, y=virus_S, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_class_with_normalized, aes(x=class, y=virus_ORF7a, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_class_with_normalized, aes(x=class, y=virus_ORF3a, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_class_with_normalized, aes(x=class, y=virus_1_261, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_class_with_normalized, aes(x=class, y=virus_ORF10, fill=class)) + geom_boxplot()
```

```{r}
ggplot(covid_class_with_normalized, aes(x=class, y=MMP9, fill=class)) + geom_boxplot()
```

# Buğra ATEŞ B1705.090046
# rm(list = ls()) 

